Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-01-04 Master Thesis/synopsis.tex
Preparing for non-finitely generatedness

We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$. The \textit{rank}\index{valuation!rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$. The \textit{rational rank}\index{valuation!rational rank} of $v:K\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
\[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
\end{definition}

@@

\begin{lemma}
Let $K$ be a finitely generated field and $v\in\operatorname{Spv}(K)$ with:
\[\operatorname{rrank}(v)+\dim k(v)= \dim(K) \]
Then $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module and $k(v)$ is a finitely generated field.
\end{lemma}

@@

\begin{lemma}
$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points. In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology.
\end{lemma}

@@

\paragraph{} Let $K$ be a finitely generated field of characteristic $0$ and dimension $2$.

@@

\begin{proof}
\begin{itemize}
\item Assume that $K$ is not algebraically closed. By the Theorem of F.K. Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are not of rank $1$. By the dimension inequality \ref{rank-dimension-inequality} and considering $\dim(K) = 2$ and $\dim k(w_i) = 0$ it follows $w_1,w_2\in\operatorname{Spv}^{(2)}_\text{rd}(K)$. Set $v_i=v(w_i)$. By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$. Again, by the Theorem of F.K. Schmidt \ref{theorem-fkschmidt}, $v_1 = v_2$. Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
\item Now assume (ii) holds.
\item The second assertion follows from the first and from [K2], §5, Theorem 1.\qedhere
\end{itemize}
\end{proof}

@@

We write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$. Let $\mathfrak{X}$ be a model of $K$. $\mathfrak{X}_i$ denotes the set of points $x\in\mathfrak{X}$ whose closure $\overline{\{x\}}$ is $i$-dimensional. For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$. By [K1] Theorem 0.8, the following map
\[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By [K1] Lemma 1.4, for a discrete valuation $v$ of $K$ the map
\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\tilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\tilde{w}}) \]
where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\tilde{w}})$. Combining these maps yields the injection:
\[ \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus concludes the proof.
\end{proof}

>>>>

We denote the valuation ring of $v\in\operatorname{Spv}(K)$ by $(\mathcal{O}_v,\mathfrak{m}_v)$ and set $U_v=\mathcal{O}_v^\times$ and $k(v)=\mathcal{O}_v/\mathfrak{m}_v$. The \textit{rank}\index{valuation!rank} of $v$ is defined as the Krull dimension of its valuation ring $\mathcal{O}_v$. The \textit{rational rank}\index{valuation!rational rank} of $v:K^\times\to\Gamma_v$ is defined as the rank of its valuation group $\Gamma_v$ as a $\mathbb{Z}$-module:
\[\operatorname{rrank}(v) = \dim_\mathbb{Q}(\Gamma_v\otimes_\mathbb{Z}\mathbb{Q}) \]
\end{definition}

@@

\begin{lemma}\label{dimension-equality}
Let $K$ be a finitely generated field and $v\in\operatorname{Spv}(K)$ with:
\[\operatorname{rrank}(v)+\dim k(v)= \dim(K) \]
Then $\Gamma_v$ is a finitely generated $\mathbb{Z}$-module and $k(v)$ is a finitely generated field.
\end{lemma}

@@

\begin{lemma}
$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. Every closed subset $A\subset\operatorname{Spv}(K)$ is the closure of its closed points.

In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology. By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.
\end{lemma}

@@

\begin{lemma}\label{lemma-spvd}
Let $K$ be a finitely generated field of dimension $2$. Then we have:
\begin{enumerate}[(i)]
\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}^{(2)}_\text{d}(K) = \operatorname{Spv}^{(2)}_\text{rd}(K)$
\item $\operatorname{Spv}(K) = \operatorname{Spv}_\text{d}(K)$
\end{enumerate}
\end{lemma}

\begin{proof}
\begin{enumerate}[(i)]
\item Let $v\in\operatorname{Spv}^{(2)}(K)\subset |\operatorname{Spv}(K)|$. By Lemma \ref{2.2-prepre} we have $\dim k(v) = 0$ and thus the equality of the dimension inequality. By Lemma \ref{dimension-equality} $\Gamma_v$ is a finitely generated abelian group of rank $2$, thus not a subgroup of $\mathbb{R}$, i.e. $\operatorname{rank}(v)> 1$. 
\item Let $v\in\operatorname{Spv}(K)$. If $\operatorname{rank}(v) = 2$, the assertion follows from Lemma \ref{rank-dimension-inequality}. If $\operatorname{rank}(v) = 0$, the value group $\Gamma_v$ is trivial, thus $\operatorname{rrank}(v) = 0$ as well. If $\operatorname{rank}(v) = 1$ and $\operatorname{rrank}(v) = 2$, we have $v\in\operatorname{Spv}^{(2)}_\text{d}(K)$ by (i), a contradiction. \qedhere
\end{enumerate}
\end{proof}

\paragraph{} Let $K$ be a (not necessarily finitely generated) field of characteristic $0$ and dimension $2$.

@@

\begin{proof}
\begin{itemize}
\item Assume that $K$ is not algebraically closed. By the Theorem of F.K. Schmidt \ref{theorem-fkschmidt}, we can see that $w_1,w_2$ are not of rank $1$. Thus they are of rank $2$ and by Lemma \ref{rank-dimension-inequality} we can see $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$. Set $v_i=v(w_i)$. By Lemma \ref{2.2-pre} $K$ is henselian with respect to both $v_1$ and $v_2$. Again, by the Theorem of F.K. Schmidt \ref{theorem-fkschmidt}, $v_1 = v_2$. Furthermore $k(v_1) = k(v_2)$ is henselian with respect to two different valuations of rank $1$, namely $w_1\mod v_1$ and $w_2\mod v_1$, making $k(v_1)$ separably closed.
\item Now assume (ii) holds.
\item The second assertion follows from the first and from [K2], §5, Theorem 1.\qedhere
\end{itemize}
\end{proof}

@@

We write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$. Let $\mathfrak{X}$ be a model of $K$. $\mathfrak{X}_i$ denotes the set of points $x\in\mathfrak{X}$ whose closure $\overline{\{x\}}$ is $i$-dimensional. For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$. By [K1] Theorem 0.8, the following map
\[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
is injective and its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By [K1] Lemma 1.4, for a discrete valuation $v$ of $K$ the map
\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\tilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\tilde{w}}) \]
where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\tilde{w}})$. Combining these maps yields the injection:
\[ \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof.
\end{proof}
